---
title: "P8105 HW 2"
author: "Haley Zylberberg"
output: github_document
---

Will first call in tidyverse and readxl libraries.

```{r}
library(tidyverse)
library(readxl)
```

## Problem 2: Clean Trash Wheel Datasets

First, import Mr. Trash Wheel dataset and clean.
```{r}
mr_trash_df = 
  read_excel("data/202207 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel", skip = 1, n_max=547) |> 
  janitor::clean_names() |>
  select (- x15, - x16, - month, - year) |>
  rename (weight_total = weight_tons, 
          volume_total = volume_cubic_yards) |>
  mutate(trash_wheel = "Mr") |> 
  relocate(trash_wheel) |>
   pivot_longer(
    plastic_bottles:sports_balls,
    names_to = "litter_type",
    values_to = "litter_amount"
  )
```

Calculate new homes powered column based on note that 1 ton garbage = 500KW of electricity and 30KW powers 1 house.
```{r}
mr_trash_df = 
mutate (mr_trash_df, homes_powered = (weight_total*500)/30)
```

Next, import Prof. Trash Wheel dataset, clean, and create new homes_powered variable.
```{r}
prof_trash_df = 
  read_excel("data/202207 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel", skip = 1, n_max =94) |> 
  janitor::clean_names() |>
  select (- month, - year) |>
  rename (weight_total = weight_tons, 
          volume_total = volume_cubic_yards) |>
  mutate(trash_wheel = "Prof") |> 
  relocate(trash_wheel) |>
   pivot_longer(
    plastic_bottles:chip_bags,
    names_to = "litter_type",
    values_to = "litter_amount"
  ) |>
mutate (homes_powered = (weight_total*500)/30)
```

Next, import Gwynnda Trash Wheel dataset, clean, and create new homes_powered variable.
```{r}
gwynnda_trash_df = 
  read_excel("data/202207 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynnda Trash Wheel", skip = 1, n_max = 106) |> 
  janitor::clean_names() |>
  select (- month, - year) |>
  rename (weight_total = weight_tons, 
          volume_total = volume_cubic_yards) |>
  mutate(trash_wheel = "Gwynnda") |> 
  relocate(trash_wheel) |>
   pivot_longer(
    plastic_bottles:plastic_bags,
    names_to = "litter_type",
    values_to = "litter_amount"
  ) |>
mutate (homes_powered = (weight_total*500)/30)
```

Combine all 3 trash wheels dataframes.
```{r}
trash_wheel_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwynnda_trash_df)
```
